$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-03 paper/paper.tex
Grammar mistake
Spell-check

% An i with a heart on top
\usepackage{pifont}
\usepackage{graphics}
\newcommand\iheart{%
    \begingroup
    \settowidth{\dimen0}{i}%
    i%
    \llap{\makebox[\dimen0][c]{%
        \kern-0.000em%
        \raisebox{1.1ex}{\scalebox{0.30}{\ding{170}}}%
    }}%
    \endgroup
}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data through communication channels.
Notably, it features \emph{channel mobility}: the channels themselves are first order constructs, and can therefore be transmitted.
The \picalc{} acts as a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
Moreover, at the state-of-the-art, the \picalc{} features a wide plethora of types: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types \cite{K07}. This makes the \picalc{} a fully-fledged tool for the modelling and verification of concurrent and distributed systems.

@@

Considering the importance of the \picalc{} and of linearity in modelling concurrent and distributed systems, in this paper we focus our research on both and go beyond.
We present for the first time a full mechanisation in Agda of a \picalc{} with linear, gradual and shared types, all under the same unified framework.
We do so by defining a specification for an algebra on multiplicities and using it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
Ultimately, we can use $\pi$ with leftovers as a unified framework for type safe distributed modelling and programming with a variety of type systems, on top of which more advanced types, theories and languages can be built.
While linearity is needed for type safety, gradual types and shared types are needed for real-world systems.

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
  \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  \autoref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- aka type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unecessary syntactic minutiae introduced by the syntax of the \picalc{}.

@@

  The typing judgments use \emph{leftover typing} to model the resource-aware \picalc{}.
  This approach adds a leftover usage context to the typing judgments.
  Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
  These technique renders explicit context splits unecessary.

@@

  \item \textbf{Fully mechanised formalisation available in Agda}:
  This work has been fully formalised in Agda and is publicly avaible \cite{Zalakain2020Agda}.

@@

\begin{nidefinition}
  Our usage algebra is a ternary relation $\op{x}{y}{z}$ that is \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  For a formal definition, refer to \autoref{fig:multiplicities} in \autoref{app:usage-algebra}.
\end{nidefinition}

@@

\begin{description}
  \item [Linear]
    The carrier is a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{one}$), and is uninhabited in every other case.
    All other properties are lifted elementwise.

@@

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Usage_{idx}^2$ to stand for a $\Usage_{idx} \times \Usage_{idx}$ tuple of input and output multiplicities, respectively.
  We use $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are lifted elementwise.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our typing rules, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
This techinque presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    The need for an extensional context splitting proof for parallel composition is eradicated.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems (\textbf{framing} (\autoref{framing}), \textbf{weakening} (\autoref{weakening}) and \textbf{strengthening} (\autoref{strengthening})) can suddenly be stated.
\end{itemize}

@@

  \begin{remark}
    We use the functionality of the monoidal relation to aliviate the user from the proof burden $\opsquared{x}{y}{z}$.
  \end{remark}

@@

Adding a new variable to the context preserves the well-typedness of a process as long as the usage annotation of that variable is preserved unchanged as a leftover.

\begin{nitheorem}
  \label{thm:weakening}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Insert an arbitrary type $t$ at position $i$ in the typing context, and an arbitrary usage annotation $x$ at position $i$ into both the input and output contexts.
  For simplicity, we overload $insert$ and use it to insert types in typing contexts and usage annotations in usage contexts.
  Increment every variable referrence greater than or equal to $i$ in process $P$.
  The resulting process is well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{remark}
  We limit ourselves to the swapping of contiguous variables $i$ and $\suc i$.
  Doing so is enough to prove subject reduction.
  The swapping of any two variables can be modeled as a sequence of contiguous variable swappings.
\end{remark}

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the most recent variable has all its $m$ multiplities exhausted, such that $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $j$ be the index of the data with $m$ multiplicities that is being received after $P$ is done, so that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute variable $\constr{zero}$ in $P$ with $\suc j$.
  As a result, the multiplicities at $\constr{zero}$ will be preserved and those at $j$ will be consumed instead, resulting in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}
  We need to proceed by induction on $P$, and to do so means going under scope restriction.
  We use the generalised version in \autoref{thm:subst-generalization} so that the variable that is being substituted is not $\constr{zero}$ but any variable $i$.
  \todo{explain arguments}
\end{proof}

@@

Work that will be done time permiting:

>>>>

% An i with a heart on top
\usepackage{pifont}
\usepackage{graphics}
\newcommand\iheart{%
    \begingroup
    \settowidth{\dimen0}{i}%
    i%
    \llap{\makebox[\dimen0][c]{%
        \color{white}%
        \raisebox{1.23ex}{\scalebox{0.40}{$\blacksquare$}}%
      }}%
    \llap{\makebox[\dimen0][c]{%
        \raisebox{1.23ex}{\scalebox{0.20}{\ding{170}}}%
    }}%
    \endgroup
}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data through communication channels.
Notably, it features \emph{channel mobility}: the channels themselves are first order constructs, and can therefore be transmitted.
The \picalc{} acts as a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{Pierce} and (more recently) Go \cite{Golang}.
Moreover, at the state-of-the-art, the \picalc{} features a wide plethora of types: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types \cite{K07}. This makes the \picalc{} a fully-fledged tool for the modeling and verification of concurrent and distributed systems.

@@

Considering the importance of the \picalc{} and of linearity in modeling concurrent and distributed systems, in this paper we focus our research on both and go beyond.
We present for the first time a full mechanisation in Agda of a \picalc{} with linear, gradual and shared types, all under the same unified framework.
We do so by defining a specification for an algebra on multiplicities and using it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
Ultimately, we can use $\pi$ with leftovers as a unified framework for type safe distributed modeling and programming with a variety of type systems, on top of which more advanced types, theories and languages can be built.
While linearity is needed for type safety, gradual types and shared types are needed for real-world systems.

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
  \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  \autoref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- aka type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.

@@

  The typing judgments use \emph{leftover typing} to model the resource-aware \picalc{}.
  This approach adds a leftover usage context to the typing judgments.
  Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
  These technique renders explicit context splits unnecessary.

@@

  \item \textbf{Fully mechanised formalisation available in Agda}:
  This work has been fully formalised in Agda and is publicly available \cite{Zalakain2020Agda}.

@@

\begin{nidefinition}
  Our usage algebra is a ternary relation $\op{x}{y}{z}$ that is \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  For a formal definition, refer to \autoref{fig:multiplicities} in \autoref{app:usage-algebra}.
\end{nidefinition}

@@

\begin{description}
  \item [Linear]
    The carrier is a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{one}$), and is uninhabited in every other case.
    All other properties are lifted element-wise.

@@

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Usage_{idx}^2$ to stand for a $\Usage_{idx} \times \Usage_{idx}$ tuple of input and output multiplicities, respectively.
  We use $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are lifted element-wise.
  Henceforth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our typing rules, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
This technique presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    The need for an extensional context splitting proof for parallel composition is eradicated.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems (\textbf{framing} (\autoref{framing}), \textbf{weakening} (\autoref{weakening}) and \textbf{strengthening} (\autoref{strengthening})) can suddenly be stated.
\end{itemize}

@@

  \begin{remark}
    We use the functionality of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  \end{remark}

@@

Inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.

\begin{nitheorem}
  \label{thm:weakening}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Insert an arbitrary type $t$ at position $i$ in the typing context, and an arbitrary usage annotation $x$ at position $i$ into both the input and output contexts.
  For simplicity, we overload $insert$ and use it to insert types in typing contexts and usage annotations in usage contexts.
  Increment every variable reference greater than or equal to $i$ in process $P$.
  The resulting process is well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{remark}
  We limit ourselves to the swapping of adjacent variables $i$ and $\suc i$.
  Doing so is enough to prove subject reduction.
  The swapping of any two variables can be modeled as a sequence of contiguous variable swappings.
\end{remark}

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the most recent variable has all its $m$ multiplicities exhausted, such that $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $j$ be the index of the data with $m$ multiplicities that is being received after $P$ is done, so that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute variable $\constr{zero}$ in $P$ with $\suc j$.
  As a result, the multiplicities at $\constr{zero}$ will be preserved and those at $j$ will be consumed instead, resulting in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}
  We need to proceed by induction on $P$, and to do so means going under scope restriction.
  We use the generalised version in \autoref{thm:subst-generalization} so that the variable that is being substituted is not $\constr{zero}$ but any variable $i$.
  \todo{explain arguments}
\end{proof}

@@

Work that will be done time permitting:
